# Working with Files

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Describe a file system, directory, and working directory
* Write paths to files or directories
* Get or set the R working directory
* Identify RDS, CSV, TSV files and functions for reading these
* Inspect the structure of a data frame
:::


## File Systems

This section is a review of how files on a computer work. You’ll need to
understand this in order to read a data set from a file, and it’s also
important for finding your saved notebooks and modules later.

Your computer's **file system** consists of **files** (chunks of data) and
**directories** (or "folders") that organize those files. For instance, the
file system on a computer shared by [Ada][ada] and [Charles][chuck], two
pioneers of computing, might look like this:

[ada]: https://en.wikipedia.org/wiki/Ada_Lovelace
[chuck]: https://en.wikipedia.org/wiki/Charles_Babbage

![
  A typical file system. Don't worry if your file system looks a bit different
  from the picture.
](/images/filesystem.png){width=50%}

File systems have a tree-like structure, with a top-level directory called the
**root directory**. On Ada and Charles' computer, the root is called `/`, which
is also what it's called on all macOS and Linux computers. On Windows, the root
is usually called `C:/`, but sometimes other letters, like `D:/`, are also used
depending on the computer's hardware.

A **path** is a list of directories that leads to a specific file or directory
on a file system (imagine giving directons to someone as they walk through the
file system). Use forward slashes `/` to separate the directories in a path.
The root directory includes a forward slash as part of its name, and doesn't
need an extra one.

For example, suppose Ada wants to write a path to the file `cats.csv`. She can
write the path like this:

```
/Users/ada/cats.csv
```

You can read this path from left-to-right as, "Starting from the root
directory, go to the `Users` directory, then from there go to the `ada`
directory, and from there go to the file `cats.csv`." Alternatively, you can
read the path from right-to-left as, "The file `cats.csv` inside of the `ada`
directory, which is inside of the `Users` directory, which is in the root
directory."

As another example, suppose Charles wants a path to the `Programs` directory.
He can write:

```
/Programs/
```

The `/` at the end of this path is reminder that `Programs` is a directory, not
a file. Charles could also write the path like this:

```
/Programs
```

This is still correct, but it's not as obvious that `Programs` is a directory.
In other words, when a path leads to a directory, including a **trailing
slash** is optional, but makes the meaning of the path clearer. Paths that lead
to files never have a trailing slash.

:::{.callout-warning}
On Windows, the components of a path are conventionally separated with
backslashes ```\``` instead of forward slashes `/`.

Regardless of the operating system, R uses forward slashes `/` to separate
components of paths. In other words, you should use paths separated by with
forward slashes in your R code, even on Windows. This is especially convenient
when you want to share code with other people, because they might use a
different operating system than you.
:::


### Absolute & Relative Paths {#sec-absolute-relative-paths}

A path that starts from the root directory, like all of the ones we've seen so
far, is called an **absolute path**. The path is "absolute" because it
unambiguously describes where a file or directory is located. The downside is
that absolute paths usually don't work well if you share your code.

For example, suppose Ada uses the path `/Programs/ada/cats.csv` to load the
`cats.csv` file in her code. If she shares her code with another pioneer of
computing, say [Gladys][gladys], who also has a copy of `cats.csv`, it might
not work. Even though Gladys has the file, she might not have it in a directory
called `ada`, and might not even have a directory called `ada` on her computer.
Because Ada used an absolute path, her code works on her own computer, but
isn't portable to others.

[gladys]: https://en.wikipedia.org/wiki/Gladys_West

On the other hand, a **relative path** is one that doesn't start from the root
directory. The path is "relative" to an unspecified starting point, which
usually depends on the context.

For instance, suppose Ada's code is saved in the file `analysis.R`, which is in
the same directory as `cats.csv` on her computer. Then instead of an absolute
path, she can use a relative path in her code:

```{.none}
cats.csv
```

The context is the location of `analysis.R`, the file that contains the code.
In other words, the starting point on Ada's computer is the `ada` directory. On
other computers, the starting point will be different, depending on where the
code is stored.

Now suppose Ada sends her corrected code in `analysis.R` to Gladys, and tells
Gladys to put it in the same directory as `cats.csv`. Since the path `cats.csv`
is relative, the code will still work on Gladys' computer, as long as the two
files are in the same directory. The name of that directory and its location in
the file system don't matter, and don't have to be the same as on Ada's
computer. Gladys can put the files in a directory `/Users/gladys/from_ada/` and
the path (and code) will still work.

Relative paths can include directories. For example, suppose that Charles wants
to write a relative path from the `Users` directory to a cool selfie he took.
Then he can write:

```{.none}
charles/cool_hair_selfie.jpg
```

You can read this path as, "Starting from wherever you are, go to the `charles`
directory, and from there go to the `cool_hair_selfie.jpg` file." In other
words, the relative path depends on the context of the code or program that
uses it.

:::{.callout-tip}
When use you paths in code, they should almost always be relative paths. This
ensures that the code is portable to other computers, which is an important
aspect of reproducibility. Another benefit is that relative paths tend to be
shorter, making your code easier to read (and write).
:::

When you write paths, there are three shortcuts you can use. These are most
useful in relative paths, but also work in absolute paths:

* `.` means the current directory.
* `..` means the directory above the current directory.
* `~` means the **home directory**. Each user has their own home directory,
  whose location depends on the operating system and their username. Home
  directories are typically `C:/Users/` on Windows, `/Users/` on macOS, and
  `/home/` on Linux.

As an example, suppose Ada wants to write a (relative) path from the `ada`
directory to Charles' cool selfie. Using these shorcuts, she can write:

```{.none}
../charles/cool_hair_selfie.jpg
```

Read this as, "Starting from wherever you are, go up one directory, then go to
the `charles` directory, and then go to the `cool_hair_selfie.jpg` file." Since
`/Users/ada` is Ada's home directory, she could also write the path as:

```{.none}
~/../charles/cool_hair_selfie.jpg
```

This path has the same effect, but the meaning is slightly different. You can
read it as "Starting from your home directory, go up one directory, then go to
the `charles` directory, and then go to the `cool_hair_selfie.jpg` file."

The `..` and `~` shortcut are frequently used and worth remembering. The `.`
shortcut is included here in case you see it in someone else's code. Since it
means the current directory, a path like `./cats.csv` is identical to
`cats.csv`, and the latter is preferable for being simpler. There are a few
specific situations where `.` is necessary, but they fall outside the scope of
this text.


### The Working Directory

@sec-absolute-relative-paths explained that relative paths have a starting
point that depends on the context where the path is used. The **working
directory** is the starting point R uses for relative paths. Think of the
working directory as the directory R is currently "at" or watching.

The function `getwd` returns the absolute path for the current working
directory, as a string. It doesn't require any arguments:

```{r}
getwd()
```

On your computer, the output from `getwd` will likely be different. This is a
very useful function for getting your bearings when you write relative paths.
If you write a relative path and it doesn't work as expected, the first thing
to do is check the working directory.

The related `setwd` function changes the working directory. It takes one
argument: a path to the new working directory. Here's an example:

```{r}
#| eval: false
setwd("..")

# Now check the working directory.
getwd()
```

:::{.callout-warning}
In your R scripts and notebooks, avoid calls to `setwd`. They make your code
more difficult to understand and to run on other computers. Use appropriate
relative paths instead.

In the R console, it's okay to occasionally use `setwd`. You might need to
change the working directory before you run some code. R's default working
directory is your home directory. In some cases, such as when you open a
project, RStudio will automatically change the working directory. However, it
doesn't always change the working directory, so `setwd` is sometimes still
necessary.
:::

Another function that's useful for dealing with the working directory and file
system is `list.files`. The `list.files` function returns the names of all of
the files and directories inside of a directory. It accepts a path to a
directory as an argument, or assumes the working directory if you don't pass a
path. For instance:

```{r}
# List files and directories in /home/.
list.files("/home/")

# List files and directories in the working directory.
list.files()
```

As usual, since you have a different computer, you're likely to see different
output if you run this code. If you call `list.files` with an invalid path or
an empty directory, the output is `character(0)`:

```{r}
list.files("/this/path/is/fake/")
```

Later on, you'll learn about what `character(0)` means more generally.


## Saving & Loading Code

:::{.callout-tip}
When you start a new project, it's a good idea to create a specific directory
for all of the project's files. If you're using R, you should also store your R
code in that directory. As you work, periodically save your code.
:::

Most of the time, you won't just write code directly into the R console.
Reproducibility and reusability are important benefits of R over
point-and-click software, and in order to realize these, you have to save your
code to your computer's hard drive.

The most common way to save R code is as an R **script** with the extension
`.R` (see @sec-reading-files for more about extensions). Editing a script is
similar to editing any other text file. You can write, delete, copy, cut, and
paste code.

In RStudio, you can create a new R script with this menu option:

```
File -> New File -> R Script
```

This will open a new pane in RStudio, like this:

![
  How RStudio typically looks after opening a new R Script.
](/images/rstudio_script.png)

The new pane is the scripts pane, which displays all of the R scripts you're
editing. Each script appears in a separate tab. In the screenshot, only one
script, the new script, is open.

Every line in an R script must be valid R code. Anything else you want to write
in the script (notes, documentation, etc.) must be placed in a comment.

Arrange your code in the order of the steps to solve the problem, even if you
write some parts before others. Comment out or delete any lines of code that
you try but ultimately decide you don't need. Make sure to save the file
periodically so that you don't lose your work. Following these guidelines will
help you stay organized and make it easier to share your code with others
later.

:::{.callout-tip}
While editing, you can run the current line in the R console by pressing
`Ctrl`-`Enter` on Windows and Linux, or `Cmd`-`Enter` on macOS. This way you
can test and correct your code as you write it.
:::


### Running Scripts

You can **source** (that is, run) an entire R script by calling the `source`
function with the path to the script as an argument. This is also what the
"Source on Save" check box refers to in RStudio. The code runs in order, only
stopping if an error occurs.

For instance, if you save the script as `my_cool_script.R`, then you can enter
`source("my_cool_script.R")` in the console to run the entire script. Pay
attention to the path---it may be different on your computer.


### Notebooks: Quarto & R Markdown

In the context of data science, a **notebook** is an interactive file that can
store a mix of code, formatted text, and images. With a notebook, you can
write, run code, and view results all in one place. Viewing and editing a
notebook requires a web browser or IDE. Some notebooks can also be converted to
static documents, such as PDFs. Comments are a good way keep notes as you
develop and run your code, but notebooks provide much more flexibility of
expression. Notebooks are a kind of [literate programming][lit].

[lit]: https://en.wikipedia.org/wiki/Literate_programming

Notebooks excel when you want to do highly interactive work and/or want to
communicate results. Use notebooks to prototype code, analyze data, refine
plots, generate documents and presentations, and practice programming. Scripts
excel when you want to reuse code (and perhaps share it as a package) or want
to run code that doesn't require much user interaction (such as time-consuming
computations you'll run on a server or high-performance computing cluster). The
remainder of this reader assumes you're using an R script rather than the R
console or a notebook, unless otherwise noted.

[Quarto][] is a popular notebook format and system for R. It also supports
Python, Julia, and other programming languages. Quarto files have the extension
`.qmd`. Quarto is based on an older notebook format, [R Markdown][rmd], which
only supports R and can't be converted to as many kinds of documents. R
Markdown is still widely used. R Markdown files have the extension `.Rmd`.

[Quarto]: https://quarto.org/
[rmd]: https://rmarkdown.rstudio.com/

:::{.callout-important}
In order to use [Quarto][], you must first download and install it. It is not
included with R or RStudio.
:::

After installing Quarto, you can create a new Quarto file in RStudio with this
menu option:

```
File -> New -> Quarto Document...
```

RStudio will prompt you to provide some details about the purpose of the file.

Notebooks are subdivided into **chunks** (or cells). You can create as many
chunks as you like, but each chunk can contain only one kind of content. You
can run a code chunk by clicking on the chunk and pressing `Ctrl`-`Enter`. The
notebook will display the result.

Markdown is a simple language you can use to add formatting to (non-code) text
in a notebook. For example, surrounding a word with asterisks, as in `Let
*sleeping* dogs lie`, makes the surrounded word italic. You can find a short,
interactive tutorial about Markdown [here][mdtutorial].

[mdtutorial]: https://www.markdowntutorial.com/

:::{.callout-note title="See Also"}
To learn more about Quarto, see the [official documentation][quarto-docs].

[quarto-docs]: https://quarto.org/docs/guide/
:::

<!--
Now that you know how file systems and paths work, you're ready to learn how to
save your R code to a file. R code is usually saved into an **R script**
(extension `.R`) or an **R Markdown file** (extension `.Rmd`). R scripts are
slightly simpler, so let's focus on those.

Editing a script is similar to editing any other text document. You can write,
delete, copy, cut, and paste text. You can also save the file to your
computer's file system. When you do, pay attention to where you save the file,
as you might need it later.
-->


## Reading Files {#sec-reading-files}

Analyzing data sets is one of the most common things to do in R. The first step
is to get R to read your data. Data sets come in a variety of file formats, and
you need to identify the format in order to tell R how to read the data.

Most of the time, you can guess the format of a file by looking at its
**extension**, the characters (usually three) after the last dot `.` in the
filename. For example, the extension `.jpg` or `.jpeg` indicates a [JPEG image
file][jpg]. Some operating systems hide extensions by default, but you can find
instructions to change this setting online by searching for "show file
extensions" and your operating system's name. The extension is just part of the
file's name, so it should be taken as a hint about the file's format rather
than a guarantee.

[jpg]: https://en.wikipedia.org/wiki/JPEG

R has built-in functions for reading a variety of formats. The R community also
provides **packages**, shareable and reusable pieces of code, to read even more
formats. You'll learn more about packages later, in @sec-packages. For now,
let's focus on data sets that can be read with R's built-in functions.

Here are several formats that are frequently used to distribute data, along
with the name of a built-in function or contributed package that can read the
format:

| Name                        | Extension  | Function or Package | Tabular?  | Text?
| :-------------------------- | :--------  | :------------------ | :-------- | :----
| Comma-separated Values      | `.csv`     | `read.csv`          | Yes       | Yes
| Tab-separated Values        | `.tsv`     | `read.delim`        | Yes       | Yes
| Fixed-width File            | `.fwf`     | `read.fwf`          | Yes       | Yes
| Microsoft Excel             | `.xlsx`    | readr package       | Yes       | No
| Microsoft Excel 1993-2007   | `.xls`     | readr package       | Yes       | No
| [Apache Arrow][arrow]       | `.feather` | arrow package       | Yes       | No
| R Data                      | `.rds`     | `readRDS`           | Sometimes | No
| R Data                      | `.rda`     | `load`              | Sometimes | No
| Plaintext                   | `.txt`     | `readLines`         | Sometimes | Yes
| Extensible Markup Language  | `.xml`     | xml2 package        | No        | Yes
| JavaScript Object Notation  | `.json`    | jsonlite package    | No        | Yes

[arrow]: https://arrow.apache.org/

A **tabular** data set is one that's structured as a table, with rows and
columns. This reader focuses on tabular data sets, since they're common in
practice and present the fewest programming challenges. Here's an example of a
tabular data set:

| Fruit  | Quantity | Price
| :----  | -------: | ----:
| apple  | 32       | 1.49
| banana | 541      | 0.79
| pear   | 10       | 1.99

A **text file** is a file that contains human-readable lines of text. You can
check this by opening the file with a text editor such as Microsoft Notepad or
macOS TextEdit. Many file formats use text in order to make the format easier
to work with.

For instance, a **comma-separated values** (CSV) file records a tabular data
using one line per row, with commas separating columns. If you store the table
above in a CSV file and open the file in a text editor, here's what you'll see:

```
Fruit,Quantity,Price
apple,32,1.49
banana,541,0.79
pear,10,1.99
```

A **binary file** is one that's not human-readable. You can't just read off the
data if you open a binary file in a text editor, but they have a number of
other advantages. Compared to text files, binary files are often faster to read
and take up less storage space (bytes).

As an example, R's built-in binary format is called RDS (which may stand for "R
data serialized"). RDS files are extremely useful for backing up work, since
they can store any kind of R object, even ones that are not tabular. You can
learn more about how to create an RDS file on the `?saveRDS` help page, and how
to read one on the `?readRDS` help page.


## Dataset: CA Least Terns {#sec-ca-least-terns}

The California least tern is a endangered subspecies of seabird that nests
along the coast of California and Mexico. The California Department of Fish and
Wildlife (CDFW) monitors least tern nesting sites across the state to estimate
breeding pairs, fledglings, and predator activity in each annual breeding
season.

![
  A California least tern. Original photo by [Mark Pavelka, U.S. Fish &
  Wildlife Service][pavelka] ([CC BY 2.0][]).
](/images/ca_least_tern_usfws_pacificsw.jpg){
  height=24em
  fig-alt="A gray bird with a white belly, black head, and orange beak sitting
  on a clutch of eggs."
}

[pavelka]: https://www.flickr.com/photos/usfws_pacificsw/5704890596/
[CC BY 2.0]: https://creativecommons.org/licenses/by/2.0/

The CDFW publishes most of the data it collects to the [California Open Data
portal][data.ca.gov]. The examples in this and subsequent chapters use a
cleaned 2000-2023 version of the California least tern data.

[data.ca.gov]: https://data.ca.gov/

:::{.callout-important}
[Click here][ca-least-tern] to download the 2000-2023 California least tern
data set.

[ca-least-tern]: https://ucdavis.box.com/s/m2w5l2ebp2rey2do5lnn38y1e3u31pin

If you haven't already, we recommend you create a directory for this workshop.
In your workshop directory, create a `data/` subdirectory. Download and save
the California least tern data set in the `data/` subdirectory.
:::

:::{.callout-note
  title="Documentation for 2000-2023 California Least Tern Data Set"
  collapse=true
}

Each row in the data set contains measurements from one year-site combination.

| Column                | Description
| --------------------: | :----------
| `year`                | Year of the breeding season
| `site_name`           | Site name
| `site_name_2013_2018` | Site name from 2013-2018
| `site_name_1988_2001` | Site name from 1988-2001
| `site_abbr`           | Abbreviated site name
| `region_3`            | Region of state: S.F. Bay, Central, or Southern (includes Ventura)
| `region_4`            | Region of state: S.F. Bay, Central, Ventura, or Southern
| `event`               | Climate events
| `bp_min`              | Reported minimum breeding pairs
| `bp_max`              | Reported maximum breeding pairs
| `fl_min`              | Reported minimum fledges
| `fl_max`              | Reported maximum fledges
| `total_nests`         | Total reported nests (maximum if a range was reported)
| `nonpred_eggs`        | Total non-predator-related mortalities of eggs
| `nonpred_chicks`      | Total non-predator-related mortalities of chicks
| `nonpred_fl`          | Total non-predator-related mortalities of fledges
| `nonpred_ad`          | Total non-predator-related mortalities of adults
| `pred_control`        | Site predator control (yes/no)
| `pred_eggs`           | Total predator-related mortalities of eggs
| `pred_chicks`         | Total predator-related mortalities of chicks
| `pred_fl`             | Total predator-related mortalities of fledges
| `pred_ad`             | Total predator-related mortalities of adults
| `pred_pefa`           | Predation by peregrine falcons (yes/no)
| `pred_coy_fox`        | Predation by coyotes or foxes (yes/no)
| `pred_meso`           | Predation by other mesocarnivores: dogs, cats, skunks, opossums, raccoons, weasels, etc. (yes/no)
| `pred_owlspp`         | Predation by owls (yes/no)
| `pred_corvid`         | Predation by corvids: ravens or crows (yes/no)
| `pred_other_raptor`   | Predation by raptors other than peregrine falcons and owls (yes/no)
| `pred_other_avian`    | Predation by birds other than raptors and corvids (yes/no)
| `pred_misc`           | Predation by other animals (yes/no)
| `total_pefa`          | Total mortalities due to peregrine falcons
| `total_coy_fox`       | Total mortalities due to coyotes and foxes
| `total_meso`          | Total mortalities due to other mesocarnivores
| `total_owlspp`        | Total mortalities due to owls
| `total_corvid`        | Total mortalities due to ravens and crows
| `total_other_raptor`  | Total mortalities due to other raptors
| `total_other_avian`   | Total mortalities due to other birds
| `total_misc`          | Total mortalities due to other animals
| `first_observed`      | Date CA least terns first observed at site
| `last_observed`       | Date CA least terns last observed at site
| `first_nest`          | Date first egg observed at site
| `first_chick`         | Date first chick observed at site
| `first_fledge`        | Date first fledge observed at site

: {.striped .hover tbl-colwidths="[25, 75]"}

The messy source data set (with more years and more columns) is available
[here][ca-least-tern-source].

[ca-least-tern-source]: https://data.ca.gov/dataset/california-least-tern-monitoring-sites-generalized-cdfw-ds3146
:::

Let's use R to read the California least tern data set. The data set is in a
file called is `2000-2023_ca_least_tern.csv`, which suggests it's a CSV file.
The function to read a CSV file is `read.csv`. The function's first and only
required argument is the path to the CSV file.

In the following code, the path to the California least tern data set is
`data/2000-2023_ca_least_tern.csv`, but it might be different for you,
depending on R's working directory and where you saved the file. We'll save the
result from the `read.csv` function in a variable called `terns`. We can use
this variable to access the data in subsequent code.

```{r}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

:::{.callout-note}
The variable name `terns` is arbitrary; you can choose something different if
you want. However, in general, it's a good habit to choose variable names that
describe the contents of the variable somehow.
:::

If you tried running the line of code above and got an error message, pay
attention to what the error message says, and remember the strategies to get
help from @sec-getting-help. The most common mistake when reading a file is
incorrectly specifying the path, so first check that you got the path right.

If the code ran without errors, it's a good idea to check that the data set
looks like what the documentation describes. When working with a new data set,
it usually isn't a good idea to print the whole thing (at least until you know
how big it is). Large data sets can take a long time to print, and the output
can be difficult to read.

Instead, use the `head` function to print only the beginning, or head, of the
data set:

```{r}
#| eval: false
head(terns)
```

```{r, echo = FALSE}
knitr::kable(head(terns))
```

If you run this code and see a similar table, then congratulations, you've read
your first data set into R! ✨

The California least terns data set is tabular---as you might have already
guessed, since it came from a CSV file. In R, it's represented by a **data
frame**, a table with rows and columns. R uses data frames to represent most
(but not all) kinds of tabular data. The `read.csv` function, which you used to
read this data, always returns a data frame.

Typically, each row in a data frame corresponds to a single subject and
is called an **observation**. Each column corresponds to a measurement of the
subject and is called a **feature** or **covariate**.

:::{.callout-note}
Sometimes people also refer to columns as “variables," but we'll try to avoid
this, because in programming contexts a variable is a name for a value (which
might not be a column).
:::

When you first read an object into R, you might not know whether it's a data
frame. One way to check is visually, by printing it (as you just did with
`head`). A better way to check is with the `class` function, which returns
information about what an object is. For a data frame, the result will always
contain `data.frame`:

```{r}
class(terns)
```

You'll learn more about classes in @sec-data-types-classes, but for now you can
use this function to identify data frames.


## Inspecting a Data Frame {#sec-data-frames}

Similar to how the `head` function shows the first six rows of a data frame,
the `tail` function shows the last six:

```{r}
tail(terns)
```

If there are lots of columns or the columns are wide, as is the case here, R
wraps the output across lines.

:::{.callout-tip}
Both `head` and `tail` accept an optional second argument that specifies the
number of rows to print:

```{r}
head(terns, 1)
```
:::

One way to get a quick idea of what your data looks like without having to skim
through all the columns and rows is by inspecting its **dimensions**. This is
the number of rows and columns in a data frame, and you can access this
information with the `dim` function:

```{r}
dim(terns)
```

So this data set has `r nrow(terns)` rows and `r ncol(terns)` columns. As an
alternative to the `dim` function, you can use the `nrow` and `ncol` functions
to get just the number of rows and number of columns, respectively.

Since the columns have names, you might also want to get just these. You can do
that with the `names` or `colnames` functions. Both return the same result:

```{r}
names(terns)
colnames(terns)
```

If the rows have names, you can get those with the `rownames` function. For
this particular data set, the rows don't have names.


### Summarizing Data

An efficient way to get a sense of what's actually in a data set is to have R
compute summary information. This works especially well for data frames, but
also applies to other data. R provides two different functions to get
summaries: `str` and `summary`.

The `str` function returns a **structural summary** of an object. This kind of
summary tells us about the structure of the data---the number of rows, the
number and names of columns, what kind of data is in each column, and some
sample values. Here's the structural summary for the least terns data set:

```{r}
str(terns)
```

This summary lists information about each column, and includes most of what you
found earlier by using several different functions separately. The summary uses
`chr` to indicate columns of text ("characters") and `int` to indicate columns
of integers.

In contrast to `str`, the `summary` function returns a **statistical summary**
of an object. This summary includes summary statistics for each column,
choosing appropriate statistics based on the kind of data in the column. For
numbers, this is generally the mean, median, and quantiles. For categories,
this is the frequencies. Other kinds of statistics are shown for other kinds of
data. Here's the statistical summary for the least terns data set:

```{r}
summary(terns)
```


### Selecting Columns

You can select an individual column from a data frame by name with `$`, the
dollar sign operator. The syntax is:

```{r}
#| eval: false
VARIABLE$COLUMN_NAME
```

For example, for the least terns data set, `terns$year` selects the `year`
column, which is the year of observation:

```{r}
terns$year
```

R provides a variety of functions to compute on columns (and other vectors of
data). For instance, what if you want to know the time period the data set
covers? You can use the `range` function to compute the minimum and maximum of
a column:

```{r}
range(terns$year)
```

So the oldest observations are from `r min(terns$year)` and the newest are from
`r max(terns$year)`, although this function and output doesn't tell us whether
there are observations for the years in between.

You can count the observations for each year with the `table` function:

```{r}
table(terns$year)
```

The `table` function is great for summarizing columns of categories, where
numerical statistics like means and standard deviations aren't defined.

On the other hand, numerical statistics work well for summarizing columns of
numbers. You can use the `mean` function to compute the mean of a column. For
instance, let's compute the mean of the `total_nests` column, which is the
total number of nests seen at a site:

```{r}
mean(terns$total_nests)
```

The result is `NA` because column is missing some values (we'll explain this in
detail in @sec-missing-values). To compute the mean with only the values that
are present, set `na.rm = TRUE` in the call to `mean`:

```{r}
mean(terns$total_nests, na.rm = TRUE)
```

You can also use the dollar sign operator to assign values to columns. For
instance, to assign `2000` to the entire `year` column:

```{r}
#| eval: false
terns$year = 2000
```

Be careful when you do this, as there is no undo. Fortunately, you haven't
saved this change to the least terns data set to your computer's hard drive
yet, so you can reload the data set to reset it:

```{r}
#| eval: false
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

In @sec-indexing, you'll learn how to select rows and individual elements from
a data frame, as well as other ways to select columns.


## Exercises


### Exercise

1. Choose a directory on your computer that you're familiar with, such as one
   you created. Determine the path to the directory, then use `list.files` to
   display its contents. Do the files displayed match what you see in your
   system's file browser?

2. What does the `all.files` parameter of `list.files` do? Give an example.


### Exercise

The `read.table` function is another function for reading tabular data. Take a
look at the help file for `read.table`. Recall that `read.csv` reads tabular
data where the values are separated by commas, and `read.delim` reads tabular
data where the values are separated by tabs.

1. What value-separator does `read.table` expect by default?
2. Is it possible to use `read.table` to read a CSV? Explain. If your answer is
   yes, show how to use `read.table` to load the least terns data set
   from @sec-ca-least-terns.
