```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Grouping

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Split data into groups and apply a function to each
* Compute group statistics with the dplyr package's `group_by` and `summarize`
  functions
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following packages:

* dplyr

@sec-packages explains how to install and load packages.
:::


## Grouping

The dplyr package's `group_by` and `summarize` functions are analogous to
`split` and purrr's map functions, but return the results in a data frame with
one row for each category. In many cases, this is more convenient than a list
or vector result.

As an example, here's how to use `group_by` and `summarize` to compute the
`total_nests` medians:

```{r}
library("dplyr")

by_region = group_by(terns, region_3)
summarize(by_region, median(total_nests, na.rm = TRUE))
```

The [dplyr documentation][dplyr] provides more details about how to use
the `group_by` and `summarize` functions.

[dplyr]: https://dplyr.tidyverse.org/




## Aggregation & Grouping

{ref}`sec-summarizing-columns` showed how to compute the mean, minimum, and
maximum values of a series. All of these functions **aggregate** the elements
of the series, reducing them to a smaller number of values (usually one).
Polars provides many different aggregation functions, which are [listed in the
documentation][pl-agg].

[pl-agg]: https://docs.pola.rs/api/python/stable/reference/expressions/aggregation.html

For example, to compute the median number of predator-related egg mortalities
across all sites and years:

```{code-cell}
terns["pred_eggs"].median()
```

With the `.select` (or `.with_columns`) method, there are two more ways you can
do this:

```{code-cell}
terns.select(pl.col("pred_eggs").median())
```

```{code-cell}
terns.select(pl.median("pred_eggs"))
```

These forms are useful if you want to compute aggregates for multiple columns
at once. For example:

```{code-cell}
terns.select(
    pl.col(pl.Float64).exclude("year").median()
)
```


### Grouping

In {ref}`sec-categorical-data`, we wrote that the identifying characteristic of
categorical features is that their values are useful for dividing observations
into groups. Now we're actually ready to do that. You can use the `.group_by`
method to group rows in a data frame by one or more columns.


To demonstrate the `.group_by` method, let's compute the median number of nests
by region for the least terns data. We'll use the `region_3` column. After
grouping with `.group_by`, you can use the `.agg` method to compute aggregates
of columns (similar to using `.select`), and Polars will compute a separate
result for each group:

```{code-cell}
terns.group_by("region_3").agg(pl.col("total_nests").median())
```

Like `.select`, the `.agg` method supports computing on multiple columns. For
example, suppose we also want the median number of breeding pairs, and want to
compute the maximum for both features as well. We'll use the `.name.suffix`
method to make sure each column in the result has a unique name:

```{code-cell}
cols = ["total_nests", "bp_max"]

terns.group_by("region_3").agg(
    pl.col(cols).median().name.suffix("_median"),
    pl.col(cols).max().name.suffix("_max")
)
```

Some aggregation functions only make sense when used with grouping. One is the
`.first` method, which returns the first row in a group. The `.first` method is
useful when all of the values in a group are the same, and you want to reduce
the data to one row per group.

For example, one way to find all year and climate event combinations present in
the least terns data is to run:

```{code-cell}
cols = ["year", "event"]

(
    terns.group_by(cols).first()
    .select(cols)
    .sort(cols)
)
```

From this result we can conclude that there is only one climate event for each
year in the data set.

For tasks like standardizing features on a per-group basis, it's necessary to
compute aggregates on groups and then map them back to observations. Rather
than using `.group_by` for this, you can use `.select` (or `.with_columns`) and
the `.over` method. For example, to compute mean total nests for each year:

```{code-cell}
terns.select(
    pl.col("year"),
    pl.col("total_nests").mean().over("year")
)
```

This approach returns a data frame with the same number of rows as the original
data frame, unlike the `.group_by` and `.agg` approach, which returns a data
frame with the same number of rows as there are groups.

:::{tip}
Use `.select` (or `.with_columns`) and `.over` if you want to compute grouped
aggregates to use in further computations on the data frame. Use `.group_by`
and `.agg` if you only want to compute grouped aggregates (for example, as a
summary, to make a visualization, or to use in further computations that are
not on the data frame).
:::


## Exercises

### Exercise

1. Compute the total number of fledglings (with `fl_min`) for each year and
   region combination.
2. Another way to present the data in {ref}`sec-example-bar-plot` is with a
   line plot. Use the data from part 1 to make this plot with points for each
   total and lines connecting the totals. Hint: find the appropriate geometries
   in the ggplot2 or Plotnine documentation.

### Exercise

1. Compute the number of sites with no egg mortalities due to predation.
2. Of those, how many had at least one fledgling?

### Exercise

1. Compute the range (minimum and maximum) of `year` for each site.
2. How many many sites have observations over the entire range of the data set
   (2000 and 2004-2023)? Hint: use `.unique` and `.len` to find the number of
   unique years for each site.


