```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Aggregation & Grouping

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Summarize columns by computing aggregates
* Summarize groups by grouping data and computing aggregates
* Explain the differences between dplyr's `summarize` and `mutate` functions
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following packages:

* dplyr

@sec-packages explains how to install and load packages.
:::

This chapter explains how to use the dplyr package to summarize columns and
groups within data frames. This is a critical skill in data analysis, and a
building block for more complicated methods.


## Summarizing Columns

@sec-working-with-files-selecting-columns showed how to compute the range and
mean of a column with R's built-in `range` and `mean` functions. These
functions summarize or **aggregate** the elements in the column, reducing them
to a smaller number of values (usually one). For example, suppose we want to
compute the median number of predator-related egg mortalities across all sites
and years in the least terns data. Then we can use the `median` function:

```{r}
median(terns$pred_eggs, na.rm = TRUE)
```

There are missing values in the column, so we have to set `na.rm = TRUE` to get
a numerical result (@sec-missing-values explains more about missing values).

The dplyr package provides functions to make it easier to summarize multiple
columns. See the [documentation][dplyr] for a complete list. Perhaps the most
useful is the `summarize` function. With `summarize`, the equivalent of the
code above is:

```{r}
library("dplyr")

summarize(terns, median(pred_eggs, na.rm = TRUE))
```

As you can see, the code is actually a bit longer! The `summarize` function is
most useful when you want to summarize multiple columns, because you can select
the columns in all of the same ways as with the `select` function. For example,
suppose we want the medians for columns `bp_min`, `bp_max`, `fl_min`, and
`fl_max`:

```{r}
summarize(
  terns,
  med_bp_min = median(bp_min, na.rm = TRUE),
  med_bp_max = median(bp_max, na.rm = TRUE),
  med_fl_min = median(fl_min, na.rm = TRUE),
  med_fl_max = median(fl_max, na.rm = TRUE)
)
```

Notice that you can set the names of the columns in the resulting data frame
through the parameter names (such as `med_bp_min`) in the call to `summarize`.


## Summarizing Groups

Categorical columns are useful for dividing observations into groups. You can
use dplyr's `group_by` function to group rows in a data frame by one or more
columns.

To demonstrate the `group_by` function, let's compute the median number of
nests by region for the least terns data. We'll use the `region_3` column.
After grouping with `group_by`, you can use the `summarize` function to compute
summaries of columns, and dplyr will compute a separate result for each group:

```{r}
by_region = group_by(terns, region_3)
summarize(by_region, median(total_nests, na.rm = TRUE))
```

The `summarize` function can compute grouped summaries for multiple columns.
For example, suppose we also want the median number of breeding pairs, and want
to compute the maximum for both features as well. We can use parameter names to
give each summary statistic a unique, easy-to-type name:

```{r}
summarize(
  by_region,
  med_total_nests = median(total_nests, na.rm = TRUE),
  med_bp_max = median(bp_max, na.rm = TRUE),
  max_total_nests = max(total_nests, na.rm = TRUE),
  max_bp_max = max(bp_max, na.rm = TRUE)
)
```

Some dplyr functions only make sense when used with grouping. One is the `n`
function, which returns the number of rows in a group. For example, one way to
find all year and climate event combinations present in the least terns data is
to run:

```{r}
summarize(
  group_by(terns, year, event),
  n()
)
```

From this result we can conclude that there is only one climate event for each
year in the dataset.


## Mutating Groups

For tasks like standardizing features on a per-group basis, it's necessary to
compute summaries on groups and then map them back to the original
observations. Use the `mutate` function for this rather than `summarize`. For
example, to compute mean total nests for each year:

```{r}
by_year = group_by(terns, year)
mutate(
  by_year,
  year,
  mean_total_nests = mean(total_nests, na.rm = TRUE),
  # Keep only the mentioned columns, to make the result easier to see.
  .keep = "used"
)
```

The `mutate` function returns a data frame with the same number of rows as the
original data frame, unlike the `summarize` function, which returns a data
frame with the same number of rows as there are groups.

:::{.callout-tip}
Use `summarize` if you want to compute grouped aggregates on their own. For
example, you might use these as a summary, to make a visualization, or to use
in further computations that are not on the original data frame.

Use `mutate` if you want to compute grouped aggregates matched to the rows of
the original data frame.
:::


## Exercises


### Exercise

1. Compute the total number of fledglings (with `fl_min`) for each year and
   region combination.
2. Another way to present the data in @sec-example-bar-plot is with a line
   plot. Use the result from part 1 to make this plot with points for each
   total and lines connecting the totals. Hint: find the appropriate geometries
   in the ggplot2 documentation.


### Exercise

1. Compute the number of sites with no egg mortalities due to predation.
2. Of those, how many had at least one fledgling?


### Exercise

1. Compute the range (minimum and maximum) of `year` for each site.
2. How many many sites have observations over the entire range of the dataset
   (2000 and 2004-2023)? Hint: use `unique` and `length` to find the number of
   unique years for each site.
