```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Grouping

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Split data into groups and apply a function to each
* Compute group statistics with the dplyr package's `group_by` and `summarize`
  functions
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following packages:

* dplyr

@sec-packages explains how to install and load packages.
:::


## Grouping

The dplyr package's `group_by` and `summarize` functions are analogous to
`split` and purrr's map functions, but return the results in a data frame with
one row for each category. In many cases, this is more convenient than a list
or vector result.

As an example, here's how to use `group_by` and `summarize` to compute the
`total_nests` medians:

```{r}
library("dplyr")

by_region = group_by(terns, region_3)
summarize(by_region, median(total_nests, na.rm = TRUE))
```

The [dplyr documentation][dplyr] provides more details about how to use
the `group_by` and `summarize` functions.

[dplyr]: https://dplyr.tidyverse.org/
