```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Grouping

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Split data into groups and apply a function to each
* Compute group statistics with the dplyr package's `group_by` and `summarize`
  functions
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following packages:

* dplyr

@sec-packages explains how to install and load packages.
:::


## Grouping

The dplyr package's `group_by` and `summarize` functions are analogous to
`split` and purrr's map functions, but return the results in a data frame with
one row for each category. In many cases, this is more convenient than a list
or vector result.

As an example, here's how to use `group_by` and `summarize` to compute the
`total_nests` medians:

```{r}
library("dplyr")

by_region = group_by(terns, region_3)
summarize(by_region, median(total_nests, na.rm = TRUE))
```

The [dplyr documentation][dplyr] provides more details about how to use
the `group_by` and `summarize` functions.

[dplyr]: https://dplyr.tidyverse.org/




## Aggregation & Grouping

@sec-summarizing-columns showed how to compute the mean, minimum, and maximum
values of a column. All of these functions **aggregate** the elements of the
series, reducing them to a smaller number of values (usually one).

For example, to compute the median number of predator-related egg mortalities
across all sites and years:

```{r}
median(terns$pred_eggs, na.rm = TRUE)
```

The dplyr package also provides a few different aggregation functions, which
are described in the [documentation][dplyr]. Perhaps the most useful is the
`summarize` function. To compute the median above with `summarize`:

```{r}
summarize(terns, median(pred_eggs, na.rm = TRUE))
```

This is useful if you want to compute aggregates for multiple columns at once.
For example:

```{r}
summarize(
  terns,
  median(bp_min, na.rm = TRUE),
  median(bp_max, na.rm = TRUE)
)
```


## Summarizing Groups

In @sec-categorical-data, we wrote that the identifying characteristic of
categorical features is that their values are useful for dividing observations
into groups. Now we're actually ready to do that. You can use dplyr's
`group_by` function to group rows in a data frame by one or more columns.


To demonstrate the `group_by` function, let's compute the median number of
nests by region for the least terns data. We'll use the `region_3` column.
After grouping with `group_by`, you can use the `summarize` function to compute
summaries of columns, and dplyr will compute a separate result for each group:

```{r}
by_region = group_by(terns, region_3)
summarize(by_region, median(total_nests, na.rm = TRUE))
```

The `summarize` function supports computing on multiple columns. For example,
suppose we also want the median number of breeding pairs, and want to compute
the maximum for both features as well. We can use parameter names to give each
summary statistic a unique, easy-to-type name:

```{r}
summarize(
  by_region,
  med_total_nests = median(total_nests, na.rm = TRUE),
  med_bp_max = median(bp_max, na.rm = TRUE),
  max_total_nests = max(total_nests, na.rm = TRUE),
  max_bp_max = max(bp_max, na.rm = TRUE)
)
```

Some aggregation functions only make sense when used with grouping. One is the
`n` function, which returns the number of rows in a group. For example, one way
to find all year and climate event combinations present in the least terns data
is to run:

```{r}
summarize(
  group_by(terns, year, event),
  n()
)
```

From this result we can conclude that there is only one climate event for each
year in the data set.


## Mutating Groups

For tasks like standardizing features on a per-group basis, it's necessary to
compute aggregates on groups and then map them back to observations. Use the
`mutate` function for this rather than `summarize`. For example, to compute
mean total nests for each year:

```{r}
by_year = group_by(terns, year)
mutate(
  by_year,
  year,
  mean_total_nests = mean(total_nests, na.rm = TRUE),
  # Keep only the mentioned columns, to make the result easier to see.
  .keep = "used"
)
```

The `mutate` function returns a data frame with the same number of rows as the
original data frame, unlike the `summarize` function, which returns a data
frame with the same number of rows as there are groups.

:::{.callout-tip}
Use `summarize` if you want to compute grouped aggregates on their own. For
example, you might use these as a summary, to make a visualization, or to use
in further computations that are not on the original data frame.

Use `mutate` if you want to compute grouped aggregates matched to the rows of
the original data frame.
:::


## Exercises


### Exercise

1. Compute the total number of fledglings (with `fl_min`) for each year and
   region combination.
2. Another way to present the data in @sec-example-bar-plot is with a line
   plot. Use the data from part 1 to make this plot with points for each total
   and lines connecting the totals. Hint: find the appropriate geometries in
   the ggplot2 or Plotnine documentation.


### Exercise

1. Compute the number of sites with no egg mortalities due to predation.
2. Of those, how many had at least one fledgling?


### Exercise

1. Compute the range (minimum and maximum) of `year` for each site.
2. How many many sites have observations over the entire range of the data set
   (2000 and 2004-2023)? Hint: use `unique` and `length` to find the number of
   unique years for each site.
