```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Indexing {#sec-indexing}

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Create vectors, including sequences
* Identify whether a function is vectorized or not
* Select the columns of a data frame
* Describe when to use `[` versus `[[`
* Filter the rows of a data frame
* Check (in)equality of values
* Negate or combine conditions with logic operators
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following packages:

* dplyr

@sec-packages explains how to install and load packages.
:::

Indexing refers to getting and setting the elements of a data structure.
Indexing is a fundamental operation in R, key to reasoning about how to solve
problems with the language. For tabular data, two kinds of indexing---selecting
columns and filtering rows---are the foundation upon which all analyses rest.
This chapter is an introduction to R's many operators and functions for
indexing.


## Vectors {#sec-vectors}

A **vector** is a collection of values. Vectors are the fundamental unit of
data in R, and you've already used them in the previous sections.

For instance, each column in a data frame is a vector. So the `site_name`
column in the California least terns data set (@sec-ca-least-terns) is a
vector. Take a look at it now. You can use `head` to avoid printing too much.
Set the second argument to `10` so that exactly 10 values are printed:

```{r}
head(terns$site_name, 10)
```

Like all vectors, this vector is **ordered**, which just means the values, or
**elements**, have specific positions. The value of the 1st element is
`r terns$site_name[[1]]`, the 2nd is `r terns$site_name[[2]]`, the 5th is
`r terns$site_name[[5]]`, and so on.

Notice that the elements of this vector are all strings. In R, the elements of
a vector must all be the same type of data (we say the elements are
**homogeneous**). A vector can contain integers, decimal numbers, strings, or
any of several other types of data, but not a mix these all at once.

The other columns in the least terns data frame are also vectors. For instance,
the `year` column is a vector of integers:

```{r}
head(terns$year)
```

Vectors can contain any number of elements, including 0 or 1 element. Unlike
mathematics, R does not distinguish between vectors and scalars (solitary
values). As far as R is concerned, a solitary value, like `3`, is a vector with
1 element.

You can check the length of a vector (and other objects) with the `length`
function:

```{r}
length(3)
length("hello")
length(terns$year)
```

Since the last of these is a column from the data frame `terns`, the length is
the same as the number of rows in `terns`.


### Creating Vectors

Sometimes you'll want to create your own vectors. You can do this by
concatenating several vectors together with the `c` function. It accepts any
number of vector arguments, and combines them into a single vector:

```{r}
c(1, 2, 19, -3)
c("hi", "hello")
c(1, 2, c(3, 4))
```

If the arguments you pass to the `c` function have different data types, R will
attempt to convert them to a common data type that preserves the information:

```{r}
c(1, "cool", 2.3)
```

@sec-implicit-coercion explains the rules for this conversion in more detail.

The colon operator `:` creates vectors that contain sequences of integers. This
is useful for creating "toy" data to test things on, and later we'll see that
it's also important in several other contexts. Here are a few different
sequences:

```{r}
1:3
-3:5
10:1
```

Beware that both endpoints are included in the sequence, even in sequences like
`1:0`, and that the difference between elements is always `-1` or `1`. If you
want more control over the generated sequence, use the `seq` function instead.


### Indexing Vectors {#sec-indexing-vectors}

You can access individual elements of a vector with the **indexing operator**
`[` (also called the square bracket operator). The syntax is:

```{r}
#| eval: false
VECTOR[INDEXES]
```

Here `INDEXES` is a vector of positions of elements you want to get or set.

For example, let's make a vector with 5 elements and get the 2nd element:

```{r}
x = c(4, 8, 3, 2, 1)
x[2]
```

Now let's get the 3rd and 1st element:

```{r}
x[c(3, 1)]
```

You can use the indexing operator together with the assignment operator to
assign elements of a vector:

```{r}
x[3] = 0
x
```

Indexing is among the most frequently used operations in R, so take some time
to try it out with few different vectors and indexes. We'll revisit indexing in
@sec-indexing to learn a lot more about it.


### Vectorization {#sec-vectorization}

Let's look at what happens if we call a mathematical function, like `sin`, on a
vector:

```{r}
x = c(1, 3, 0, pi)
sin(x)
```

This gives us the same result as if we had called the function separately on
each element. That is, the result is the same as:

```{r}
c(sin(1), sin(3), sin(0), sin(pi))
```

Of course, the first version is much easier to type.

Functions that take a vector argument and get applied element-by-element like
this are said to be **vectorized**. Most functions in R are vectorized,
especially math functions. Some examples include `sin`, `cos`, `tan`, `log`,
`exp`, and `sqrt`.

Functions that are not vectorized tend to be ones that combine or aggregate
values in some way. For instance, the `sum`, `mean`, `median`, `length`, and
`class` functions are not vectorized.

A function can be vectorized across multiple arguments. This is easiest to
understand in terms of the arithmetic operators. Let's see what happens if we
add two vectors together:

```{r}
x = c(1, 2, 3, 4)
y = c(-1, 7, 10, -10)
x + y
```

The elements are paired up and added according to their positions. The other
arithmetic operators are also vectorized:

```{r}
x - y
x * y
x / y
```

### Recycling {#sec-recycling}

When a function is vectorized across multiple arguments, what happens if the
vectors have different lengths? Whenever you think of a question like this as
you're learning R, the best way to find out is to create some toy data and test
it yourself. Let's try that now:

```{r}
x = c(1, 2, 3, 4)
y = c(-1, 1)
x + y
```

The elements of the shorter vector are **recycled** to match the length of the
longer vector. That is, after the second element, the elements of `y` are
repeated to make a vector with the same length as `x` (because `x` is longer),
and then vectorized addition is carried out as usual.

Here's what that looks like written down:

```{.none}
   1  2  3  4
+ -1  1 -1  1
  -----------
   0  3  2  5
```

If the length of the longer vector is not a multiple of the length of the
shorter vector, R issues a warning, but still returns the result. The warning
as meant as a reminder, because unintended recycling is a common source of
bugs:

```{r}
x = c(1, 2, 3, 4, 5)
y = c(-1, 1)
x + y
```

Recycling might seem strange at first, but it's convenient if you want to use a
specific value (or pattern of values) with a vector. For instance, suppose you
want to multiply all the elements of a vector by `2`. Recycling makes this
easy:

```{r}
2 * c(1, 2, 3)
```

When you use recycling, most of the time one of the arguments will be a scalar
like this.


## More Than One Dimension

@sec-indexing-vectors introduced the indexing (or square bracket) operator
`[` as a way to get and set the elements of a vector. You can also use the
indexing operator with many other data structures, such as data frames, lists,
matrices, and arrays.

Consider data frames. A data frame is two-dimensional: it has rows and columns.
In order to describe the exact position of an element in a data frame, we need
to specify both a row index and a column index. You can still use the indexing
operator `[` to get and set elements, but now it takes two arguments:

```{r}
#| eval: false
DATA[ROW_INDEXES, COLUMN_INDEXES]
```

For example, to get the value at row 1 and column 2 in the California least
terns data (@sec-ca-least-terns), you can write:

```{r}
terns[1, 2]
```

If you swap the order of the indexes, you get the value at row 2 and column 1
instead:

```{r}
terns[2, 1]
```

As with indexing vectors, when indexing a data frame you can also:

* Use a vector as an index to get multiple elements.
* Leave an index blank to get everything in that dimension.
* Use negative indexes to omit rows or columns.
* Combine indexing with assignment to set elements.


Let's look at examples of the first two. Suppose we want to get elements from
multiple rows and columns. Here's the code to get elements from the first 3
rows in columns 2 and 7 of the least terns data:

```{r}
terns[1:3, c(2, 7)]
```

Now suppose we want to get the elements from rows 3, 5, and 7 in all columns.
Then you can write:

```{r}
terns[c(3, 5, 7), ]
```

By leaving the column index blank, you're telling R to select all of the
columns. Likewise, you can leave the row index blank to select all of the rows.

:::{.callout-note}
You can also use the indexing operator `[` with data structures that have more
than two dimensons. Maybe you can guess the syntax:

```{r}
#| eval: false
DATA[DIM1_INDEXES, DIM2_INDEXES, DIM3_INDEXES, ...]
```

You won't encounter any data structures with more than two dimensions in this
course, but you might someday.
:::


## Selecting Columns

In order to work with tabular data, you need to know how to **select** (or get)
columns. Most of the time, it's best to select columns by name. Doing so makes
code easier to understand, since the reader doesn't have to memorize the
positions of the columns. It can also make code more reliable, since the code
will still work even if the positions of the columns change.

This section explains several different ways to select columns and how to
choose which one to use in any given situation. Generally, the first thing to
think about is whether you want to select just one column or multiple columns.


### One Column {#sec-one-column}

The dollar sign operator `$`, introduced in @sec-inspecting-a-data-frame, is
one way to select a single column from a data frame. For example, to select the
`total_nests` column from the least terns dataset:

```{r}
terns$total_nests
```

The dollar sign operator always selects the column by name (not position).

Another way to select a single column is with the **extraction operator** `[[`
(also called the double square bracket operator). The column name must be
quoted and goes inside the brackets. For instance:

```{r}
terns[["total_nests"]]
```

The double square bracket operator is helpful if you want to select a column
with a name you've stored in a variable:

```{r}
my_column = "site_name"
terns[[my_column]]
```

The double square bracket operator is also important for getting and setting
elements of lists (@sec-lists). Unlike the dollar sign operator, it can also
get columns by position:

```{r}
terns[[1]]
```

Most of the time, you should use the dollar sign operator when you want to get
a single column.

:::{.callout-tip}
Quotes are optional with the dollar sign operator. So another way to get the
`total_nests` column is:

```{r}
#| eval: false
terns$"total_nests"
```

Use quotes if you want to get a column whose name contains spaces or other
special characters. For all other cases, quotes are not necessary.
:::


### Multiple Columns

The dollar sign operator `$` and double square bracket operator `[[` can only
get one column of a data frame at a time, and they always return columns as
vectors. This is useful when we want to compute on just one column, because
most functions (especially math and statistics functions) require vector
arguments.

Sometimes we'll want to get multiple columns from a data frame. Imagine a data
frame with 300 columns, but only 5 that are relevant to your research question.
Getting rid of the extra columns makes the data easier to inspect and might
make some computations faster. When you select multiple columns from a data
frame, the result is also a data frame.

One way to select multiple columns is with the indexing operator `[`. If you
use the indexing operator on a data frame and only provide one index, R will
select columns. For example, to select the `site_name` and `year` columns from
the least terns dataset:

```{r}
terns[c("site_name", "year")]
```

:::{.callout-warning}
In R, omitting an argument is different from leaving the argument blank. So
these two expressions do different things despite looking similar:

```{r}
#| eval: false
terns[c("site_name", "year")]
```

```{r}
#| eval: false
terns[c("site_name", "year"), ]
```

The first selects the `site_name` and `year` columns. The second selects the
`site_name` and `year` rows (which causes an error, since the least terns
dataset doesn't have row names).

There's one more possibility:

```{r}
#| eval: false
terns[, c("site_name", "year")]
```

Like the first expression, this selects the `site_name` and `year` columns. You
can write the code either way. Some people prefer the first expression because
it's shorter, while others prefer this expression because it makes it clear
that `terns` is two-dimensional.
:::

You can select whatever combination of columns you want with the indexing
operator, but it's somewhat inconvenient if you need to select a lot of
columns, because you typically have to type out all of their names. For
example, suppose we want to select all of the columns in the least terns
dataset whose name starts with `nonpred_` (these are mortalities caused by
something other than predation).

The [dplyr][] package is a collection of functions that make working with data
frames more convenient. Install and load the package if you want to follow
along with this example:

[dplyr]: https://dplyr.tidyverse.org/

```{r}
# install.packages("dplyr")
library("dplyr")
```

Remember that you only need to install each package once.

The dplyr package's `select` function selects columns from a data frame. Unlike
most of R's built-in functions, the functions in dplyr don't require quotes
around column names. To select the `site_name` and `year` columns from the
least terns dataset, you can write:

```{r}
select(terns, site_name, year)
```

The advantage of `select` over the indexing operator `[` is that it provides
concise ways to a lot of columns. For instance, to select all of the columns
whose names begin with `nonpred_`, you can write:

```{r}
select(terns, starts_with("nonpred_"))
```

You can select the same columns with the indexing operator, but you'd either
have to type out all of their names or use a more complicated expression.

Use the indexing operator when you only need to select a few columns. Use
dplyr's `select` function when you need to select lots of columns that follow
some kind of pattern. The [dplyr documentation][dplyr] provides more details
about how to use the `select` function.


## Comparison Operators {#sec-comparisons}

Besides arithmetic, you can also use R to compare values. Programming tasks
often involve comparing values. Use **comparison operators** to do so:

* `<` for "less than"
* `>` for "greater than"
* `<=` for "less than or equal to"
* `>=` for "greater than or equal to"
* `==` for "equal to"
* `!=` for "not equal to"

Notice that the "equal to" operator is two equal signs. This is to distinguish
it from the assignment operator `=`.

Let's look at a few examples:
```{r}
1.5 < 3
"a" > "b"
pi == 3.14
"hi" == 'hi'
```

When you make a comparison, R returns a **logical** value, `TRUE` or `FALSE`,
to indicate the result. Logical values are not the same as strings, so they are
not quoted.

Logical values are values, so you can use them in other computations. For
example:

```{r}
TRUE
TRUE == FALSE
```

@sec-logic describes more ways to use and combine logical values.

:::{.callout-warning}
Some of R's equality operators return `TRUE` even when comparing two different
types of data:

```{r}
"1" == 1
"TRUE" <= TRUE
"FALSE" <= TRUE
```

@sec-implicit-coercion explains why this happens, and
@sec-more-about-comparisons explains several other ways to compare values.
:::


## Filtering Rows

In addition to selecting columns, you also need to know how to **filter** rows
in order to work with tabular data. In this context, to "filter" is to get a
subset of rows that satisfy a condition. As a motivating example, suppose we
want to get all of the rows in the least terns dataset that correspond to the
Los Angeles Harbor (`LA HARBOR`) site.

The indexing operator `[` is one way to filter the rows of a data frame. The
first step is to translate the condition on the rows into code. To test whether
rows correspond to Los Angeles Harbor, we can write:

```{r}
terns$site_name == "LA HARBOR"
```

This returns a vector of logical (`TRUE` and `FALSE`) values. Now that we have
the code for the condition, we can use indexing to get rows where the condition
is `TRUE`:

```{r}
terns[terns$site_name == "LA HARBOR", ]
```

@sec-logic describes various operators and functions you can use to write more
complicated conditions.

Notice that in the condition, we had to write `terns$site_name` rather than
just the column name `site_name`. We have to specify that we mean a column in
`terns` because the indexing operator doesn't make any assumptions about what
is or isn't a column name. But most of the time, we filter rows by testing
conditions on the columns. Typing the name of the data frame (for example,
`terns$`) in front of every column name makes writing complicated conditions
tedious.

The dplyr package's `filter` function is another way to filter the rows of a
data frame. By assuming that names refer to column names, `filter` makes
writing conditions less tedious. The first argument to `filter` is the data
frame to filter, while the second is the condition. Here's how to get the Los
Angeles Harbor rows from the least terns dataset:

```{r}
filter(terns, site_name == "LA HARBOR")
```

You can pass additional conditions to `filter` as additional arguments. All of
them must be satisfied (`TRUE`) for a row to be kept.

:::{.callout-note}
R's built-in `subset` function is similar to and might have inspired dplyr's
`filter` function. Here's how to use it to get all of the Los Angeles Harbor
rows:

```{r}
#| eval: false
subset(terns, site_name == "LA HARBOR")
```

The `subset` function is a good alternative if you prefer not to use dplyr.
:::

Both the indexing operator and dplyr's `filter` function are good ways to
filter rows in a data frame. The indexing operator has the advantage that you
can simultaneously use it to select columns by filling in the second argument.
On the other hand, the `filter` function makes complicated conditions easier to
type and read---this might seem like a small advantage, but code clarity is
important.


## Logic Operators {#sec-logic}

All of the conditions we've seen so far have been written in terms of a single
test. If you want to use more sophisticated conditions, R provides operators to
negate and combine logical vectors. These operators are useful for working with
logical vectors even outside the context of indexing.


### Negation

The **NOT operator** `!` converts `TRUE` to `FALSE` and `FALSE` to `TRUE`:
```{r}
x = c(TRUE, FALSE, TRUE, TRUE, NA)
x
!x
```

You can use `!` with a condition:
```{r}
y = c("hi", "hello")
!(y == "hi")
```

The NOT operator is vectorized.


### Combinations

R also has operators for combining logical values.

The **AND operator** `&` returns `TRUE` only when both arguments are `TRUE`.
Here are some examples:

```{r}
FALSE & FALSE
TRUE & FALSE
FALSE & TRUE
TRUE & TRUE
c(TRUE, FALSE, TRUE) & c(TRUE, TRUE, FALSE)
```

The **OR operator** `|` returns `TRUE` when at least one argument is `TRUE`.
Let's see some examples:

```{r}
FALSE | FALSE
TRUE | FALSE
FALSE | TRUE
TRUE | TRUE
c(TRUE, FALSE) | c(TRUE, TRUE)
```

The combination operators are vectorized.

:::{.callout-warning}
Be careful: everyday English is less precise than logic. You might say:

> I want all subjects with age over 50 and all subjects that like cats.

But in logic this means:

`(subject age over 50) OR (subject likes cats)`

So think carefully about whether you need both conditions to be true (AND) or
at least one (OR).
:::

:::{.callout-note title="Short-circuit Operators" collapse="true"}
The second argument is irrelevant in some conditions:

* `FALSE &` is always `FALSE`
* `TRUE |` is always `TRUE`

Now imagine you have `FALSE & long_computation()`. You can save time by
skipping `long_computation()`. A **short-circuit operator** does exactly that.

R has two short-circuit operators:

* `&&` is a short-circuited `&`
* `||` is a short-circuited `|`

These operators only evaluate the second argument if it is necessary to
determine the result. Here are some of these:

```{r}
TRUE && FALSE
TRUE && TRUE
TRUE || TRUE
```

The short-circuit operators are not vectorized---they only accept length-1
arguments:

```{r, error = TRUE}
c(TRUE, FALSE) && c(TRUE, TRUE)
```

Because of this, you can't use short-circuit operators for indexing. Their main
use is in writing conditions for if-expressions, which we'll learn about later
on.

Prior to R 4.3.0, short-circuit operators didn't raise an error for inputs with
length greater than 1 (and thus were a common source of bugs).
:::

:::{.callout-note title="Exclusive OR" collapse="true"}
Rarely, you might want _exactly one_ condition to be true. The **XOR (eXclusive
OR) function** `xor()` returns `TRUE` when exactly one argument is `TRUE`. For
example:

```{r}
xor(FALSE, FALSE)
xor(TRUE, FALSE)
xor(TRUE, TRUE)
```
:::


## Exercises


### Exercise

The `rep` function is another way to create a vector. Read the help file for
the `rep` function.

1. What does the `rep` function do to create a vector? Give an example.
2. The `rep` function has parameters `times` and `each`. What does each do, and
   how do they differ? Give examples for both.
3. Can you set both of `times` and `each` in a single call to `rep`? If the
   function raises an error, explain what the error message means. If the
   function returns a result, explain how the result corresponds to the
   arguments you chose.


### Exercise

1.  Create a new data frame from the least terns data with the following
    characteristics:
      * Each entry's year is between 2010 and 2019 (inclusive).
      * Each entry reports at least 100 breeding pairs.
      * The columns are `year`, `site_name`, `bp_min`, `bp_max`,
        `total_nests`.

    Use this data frame for the remaining questions.

2.  Count the number of entries for each site. How many sites have at least 100
    breeding pairs across all 10 years?

3.  Which site-year combination has the highest number of nests?
