```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Visualization {#sec-data-visualization}

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Describe the grammar of graphics
* Describe the three required layers in every ggplot2 plot
* Make a plot
* Save a plot to an image file
* Explain some principles of visualization design
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following packages:

* ggplot2

@sec-packages explains how to install and load packages.
:::

In this chapter, you'll learn how to efficiently explore and summarize data
with visualizations.


## The Visualization Landscape

There are many ways to create visualizations in R, because visualization is a
highly effective way to summarize and gain insights about data. Tools for
creating static visualizations include:

* R's built-in base and graphics packages provide many data visualization
  functions. For instance, the `plot` function creates scatter plots and line
  plots, while the `barplot`, `boxplot`, `histogram`, and `smoothScatter`
  functions are named after the plots they create. With these functions, adding
  more details to a plot is relatively straightforward (for example, use the
  `lines` function to add lines). There's also no need to install any packages
  to use them. A disadvantage is that they all have slightly different
  parameters, which can make them difficult to learn and remember.
* The [ggplot2][] package uses the grammar of graphics, a convenient way to
  describe visualizations in terms of layers. The package provides a consistent
  interface, detailed [documentation][ggplot2], and a
  [cheatsheet][ggplot2-cheat], all of which are helpful for learning. As R's
  most-downloaded package, ggplot2 is so popular that people have created
  equivalents for the Python and Julia programming languages. The package is
  part of the [Tidyverse][tidy] (@sec-the-tidyverse).
* The [lattice][] package is another package that aims to provide rich support
  for multivariate and grouped data. While lattice creates nice-looking plots
  by default, customizing them usually requires writing a lot of boilerplate
  code.

[ggplot2]: https://ggplot2.tidyverse.org/
[ggplot2-cheat]: https://github.com/rstudio/cheatsheets/blob/master/data-visualization.pdf
[tidy]: https://www.tidyverse.org/
[lattice]: https://cran.r-project.org/web/packages/lattice/

Many packages extend these systems further. There are also many packages for
creating dynamic visualizations, which react to user input or change over time,
that we do not cover here.

:::{.callout-warning}
R's built-in data visualization functions, ggplot2, and lattice are not
interoperable! Consequently, it's best to choose one system to use exclusively.

We recommend ggplot2 and will focus on it for the rest of this chapter.
:::


## The Grammar of Graphics

The "gg" in ggplot2 stands for **grammar of graphics**. The idea of a grammar
of graphics is that visualizations can be built up in layers. In ggplot2, you
can add layers to a plot with plus (`+`). There are three layers that every
plot must have are:

Layer      | Functions | Description
---------- | --------- | -----------
data       | `ggplot2` | Sets the data used to make the plot.
geometry   | `geom_*`  | Sets the geometry (points, lines, etc.) of the plot.
aesthetics | `aes`     | Sets the relationship between data and geometries.

Each layer corresponds to one or more functions. The geometry layer consists of
multiple functions; the asterisk (`*`) in the table is a wildcard.

There are also several optional layers. Here are a few:

Layer       | Functions      | Description
----------- | -------------- | -----------
scale       | `scale_*`      | Sets palettes (color, shape, etc.) and labels
theme       | `theme_*`      | Sets appearance of non-data elements
facets      | `facet_*`      | Makes side-by-side plots
annotation  | `annotation_*` | Adds manually-positioned annotations
guides      | `guide_*`      | Sets axes and legends
coordinate  | `coord_*`      | Sets coordinate systems (Cartesian, logarithmic, polar)

The ggplot2 [documentation][ggplot2] provides a complete list of all layers and
their corresponding functions.

:::{.callout-tip}
The grammar of graphics breaks the elements of statistical graphics into parts
in an analogy to human language grammars. Knowing how to put together subject
nouns, object nouns, verbs, and adjectives allows you to construct sentences
that express meaning. Similarly, the grammar of graphics is a collection of
layers and the rules for putting them together to graphically express the
meaning of your data.
:::


## Making a Plot

Let's visualize the California least terns dataset from @sec-ca-least-terns to
see how the grammar of graphics works in practice. But what kind of plot should
we make? It depends on what we want to know about the dataset!

Suppose we want to understand the relationship between the number of breeding
pairs and the total number of nests at each site, and whether this relationship
is affected by climate events. One way to show the relationship between two
numerical features like these is to make a scatter plot.

Before we can make the plot, we need to load the ggplot2 package. Install the
package if you haven't already, and then load it:

```{r}
# install.packages("ggplot2")
library("ggplot2")
```

### Layer 1: Data

The **data layer** determines the dataset(s) used to make the plot.

ggplot2 and most other Tidyverse packages are designed to work with **tidy**
data, which means:

1. Each feature has its own column.
2. Each observation has its own row.
3. Each value has its own cell.

These rules ensure data are easy to read visually and access with indexing. The
least terns dataset satisfies all of these rules.

:::{.callout-note title="See also"}
All of the datasets we use in this reader are tidy. To learn how to tidy an
untidy dataset, see the [Untidy & Relational Data][dl-r-tidy] chapter of
DataLab's Intermediate R workshop reader.

[dl-r-tidy]: https://ucdavisdatalab.github.io/workshop_intermediate_r/tidy-relational-data.html
:::

To set up the data layer, call the `ggplot` function on a data frame:
```{r}
ggplot(terns)
```

This returns a blank plot. We still need to add a few more layers.


### Layer 2: Geometry

The **geometry layer** determines the shape or appearance of the visual
elements of the plot. In other words, the geometry layer determines what kind
of plot to make: one with points, lines, boxes, or something else.

There are many different geometries available in ggplot2. The package provides
a function for each geometry, always prefixed with `geom_`.

To add a geometry layer to the plot, choose the `geom_` function you want and
add it to the plot with the `+` operator. We'll use `geom_point`, which makes a
scatter plot (a plot with points):

```{r, error=TRUE, fig.show="hide"}
ggplot(terns) + geom_point()
```

This returns an error message that we're missing aesthetics `x` and `y`. We'll
learn more about aesthetics in the next section, but this error message is
especially helpful: it tells us exactly what we're missing. When you use a
geometry you're unfamiliar with, it can be helpful to run the code for just the
data and geometry layer like this, to see exactly which aesthetics need to be
set.

As we'll see later, it's possible to add multiple geometries to a plot.


### Layer 3: Aesthetics

The **aesthetic layer** determines the relationship between the data and the
geometry. Use the aesthetic layer to map features in the data to aesthetics
(visual elements) of the geometry.

The `aes` function creates an aesthetic layer. The syntax is:

```{r}
#| eval: false
aes(AESTHETIC = FEATURE, ...)
```

The names of the aesthetics depend on the geometry, but some common ones are
`x`, `y`, `color`, `fill`, `shape`, and `size`. There is more information about
and examples of aesthetic names in the documentation.


For the scatter plot of breeding pairs against total nests, we'll put `bp_min`
on the x-axis and `total_nests` on the y-axis. Below, we set both of these
aesthetics. We also enclose all of the code for the plot in parentheses `()` so
that we can put the code for each layer on a separate line, which makes the
layers easier to distinguish:

```{r}
ggplot(terns) +
  aes(x = bp_min, y = total_nests) +
  geom_point()
```

:::{.callout-important}
In the `aes` function, column names are never quoted.
:::

:::{.callout-note collapse="true" title="Note: The Old Aesthetic Layer Syntax"}
In older versions of ggplot2, you must pass the aesthetic layer as the second
argument of the `ggplot` function rather than using `+` to add it to the plot.
This syntax is still widely used:

```{r}
ggplot(terns, aes(x = bp_min, y = total_nests)) +
  geom_point()
```
:::

At this point, we've supplied all three layers necessary to make a plot: data,
geometry, and aesthetics. The plot shows what looks like a linear relationship
between number of breeding pairs and total nests. To refine the plot, you can
add more layers and/or set parameters on the layers you have.

Let's add another aesthetic to the plot: we'll make the color and shape of each
point correspond to `event`, the climate event for each observation:

```{r}
ggplot(terns) +
  aes(x = bp_min, y = total_nests, color = event, shape = event) +
  geom_point()
```

Using color and shape for the same feature is redundant, but ensures that the
plot is accessible to colorblind people.


#### Additional Geometries {-}

Each observation in the least terns data corresponds to a specific year and
site. What if we label the points with their years? You can add text labels to
a plot with `geom_text`. The required aesthetic for this geometry is `label`:

```{r}
ggplot(terns) +
  aes(
    x = bp_min, y = total_nests,
    color = event, shape = event,
    label = year
  ) +
  geom_point() +
  geom_text()
```

The labels make the plot more difficult to read and probably would even if we
made them smaller, because there are so many points on the plot. Making a
high-quality visualization is typically a process of drafting and revising,
similar to writing a high-quality essay. In this example, adding year labels to
the plot doesn't work well, so we'll backtrack and leave them off of the plot.
If accounting for year was critical to our research question, we could do it in
other ways, such as by making separate plots for each year.

#### Per-geometry Aesthetics {-}

Before we remove the labels, let's use them to demonstrate an important point
about using multiple geometry and aesthetic layers: when you add an aesthetic
layer to a plot, it applies to the entire plot. You can also set an aesthetic
layer for an individual geometry by passing the layer as the first argument in
the `geom_` function. Here's the same plot as above, but with the `color`
aesthetic only set for the labels:

```{r}
ggplot(terns) +
  aes(
    x = bp_min, y = total_nests,
    shape = event,
    label = year
  ) +
  geom_point() +
  geom_text(aes(color = event))
```

Notice that the points are no longer color-coded. Where you put aesthetic
layers matters.


#### Constant Aesthetics {-}

If you want to set an aesthetic to a constant value, rather than one that's
data dependent, do so in the geometry layer rather than the aesthetic layer.

For instance, suppose we want to make all of the points blue and use only point
shape to indicate climate events:

```{r}
ggplot(terns) +
  aes(
    x = bp_min, y = total_nests,
    shape = event
  ) +
  geom_point(color = "blue")
```

If you set an aesthetic to a constant value inside of the aesthetic layer, the
results you get might not be what you expect:

```{r}
ggplot(terns) +
  aes(
    x = bp_min, y = total_nests,
    color = "blue", shape = event,
    label = year
  ) +
  geom_point() +
  geom_text()
```


### Layer 4: Scales

The **scales layer** controls the title, axis labels, and axis scales of the
plot. Most of the functions in the scales layer are prefixed with `scale_`, but
not all of them.

The `labs` function is especially important, because it's used to set the title
and axis labels. Visualizations should generally have a title and axis labels,
to aid the viewer:

```{r}
ggplot(terns) +
  aes(
    x = bp_min, y = total_nests,
    color = event, shape = event
  ) +
  geom_point() +
  labs(
    x = "Minimum Reported Breeding Pairs",
    y = "Total Nests",
    color = "Climate Event", shape = "Climate Event",
    title = "California Least Terns: Breeding Pairs vs. Nests"
  )
```

Notice that to set the title for a legend with `labs`, you can set the
parameters of the same names as the corresponding aesthetics. While our plot is
still far from perfect---some of the points are hard to see because of how many
there are---it's now good enough to provide some insight into the relationship
between number of breeding pairs and nests.


## Saving a Plot

You can use the `ggsave` function to save a plot you've assigned to a variable
or the most recent plot you created (with no argument to `ggsave`):

```{r}
#| eval: false
ggsave("myplot.png")
```

The file format is selected automatically based on the extension. Common
formats include PNG, TIFF, SVG, and PDF.

:::{tip}
PNG and SVG are good choices for sharing visualizations online, while TIFF and
PDF are good choices for print. Many journals require that visualizations be in
TIFF format.
:::


:::{.callout-note collapse="true" title="Note: R Plot Devices"}
You can also save a plot with one of R's "plot device" functions. The steps
are:

1. Call a plot device function: `png`, `jpeg`, `pdf`, `bmp`, `tiff`, or `svg`.
2. Run your code to make the plot.
3. Call `dev.off` to indicate that you're done plotting.

This strategy works with any of R's graphics systems (not just ggplot2).

Here's an example:
```{r}
#| eval: false
# Run these lines in the console, not the notebook!
jpeg("myplot.jpeg")
ggplot(terns) +
  aes(
    x = bp_min, y = total_nests,
    color = event, shape = event
  ) +
  geom_point() +
  labs(
    x = "Minimum Reported Breeding Pairs",
    y = "Total Nests",
    color = "Climate Event", shape = "Climate Event",
    title = "California Least Terns: Breeding Pairs vs. Nests"
  )
dev.off()
```
:::


## Example: A Bar Plot {#sec-example-bar-plot}

Suppose we want to visualize how many fledglings there are each year, further
broken down by region. A bar plot is one appropriate way to represent this
visually.

The geometry for a bar plot is `geom_bar`. Since bar plots are mainly used to
display frequencies, by default the `geom_bar` function counts the number of
observations in each category on the x-axis and displays these counts on the
y-axis. You can make `geom_bar` display values from a column on the y-axis by
setting the `weight` aesthetic:

```{r}
ggplot(terns) +
  aes(x = year, weight = fl_min, fill = region_3) +
  geom_bar()
```

:::{.callout-note collapse="true" title="Note: Setting the Statistics Layer"}
Every geometry layer has a corresponding statistics layer, which transforms
feature values into quantities to plot. For many geometries, the default
statistics layer is the only one that makes sense.

Bar plots are an exception. The default statistics layer is `stat_count`, which
counts observations. If you already have counts (or just want to display some
quantities as bars), you need `stat_identity` (or the `weight` aesthetic
described above). Here's one way to change the statistics layer:

```{r}
ggplot(terns) +
  aes(x = year, y = fl_min, fill = region_3) +
  geom_bar(stat = "identity")
```

This produces the same plot as setting `weight` and using the default
statistics layer `stat_count`.
:::

The plot reveals that there are a few extraneous categories in the `region_3`
column: `ARIZONA`, `KINGS`, and `SACRAMENTO`. These might or might not be
erroneous---and it would be good to investigate---but they don't add anything
to this plot, so let's exclude them.

Let's also change the **color map**, the palette of colors used for the
categories. These are both properties of the scale layer for the `fill`
aesthetic, so we'll use a `scale_fill_` function. In particular, we'll use the
"viridis" color map, and since the fill color corresponds to categorical
(discrete) data, we'll use `scale_fill_viridis_d`. We'll also add labels:

```{r}
terms_to_keep = c("S.F._BAY", "CENTRAL", "SOUTHERN")
terns_filtered = terns[terns$region_3 %in% terms_to_keep, ]

ggplot(terns_filtered) +
  aes(x = year, weight = fl_min, fill = region_3) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(
    title = "California Least Terns: Fledglings",
    x = "Year",
    y = "Minimum Reported Fledglings",
    fill = "Region"
  )
```

You can read more about the viridis color map in ggplot2's [documentation for
this function][gg-viridis]. The plot reveals that the dataset is missing
2001-2003 and that overall, fledgling counts seem to be declining in recent
years.

[gg-viridis]: https://ggplot2.tidyverse.org/reference/scale_viridis.html

:::{.callout-tip}
The setting `position = "dodge"` instructs `geom_bar` to put the bars
side-by-side rather than stacking them.
:::


## Visualization Design

Designing high-quality visualizations goes beyond just mastering which R
functions to call. You also need to think carefully about what kind of data you
have and what message you want to convey. This section provides a few
guidelines.

The first step in data visualization is choosing an appropriate kind of plot.
Here are some suggestions (not rules):

| Feature 1   | Feature 2   | Plot
| :---------- |:----------- |:----
| categorical |             | bar, dot
| categorical | categorical | bar, dot, mosaic
| numerical   |             | box, density, histogram
| numerical   | categorical | box, density, ridge
| numerical   | numerical   | line, scatter, smooth scatter

If you want to add a:

* 3rd numerical feature, use it to change point/line sizes.
* 3rd categorical feature, use it to change point/line styles.
* 4th categorical feature, use side-by-side plots.

Once you've selected a plot, here are some rules you should almost always
follow:

* Always add a title and axis labels. These should be in plain English, not
  variable names!
* Specify units after the axis label if the axis has units. For instance,
  "Height (ft)".
* Don't forget that many people are colorblind! Also, plots are often printed
  in black and white. Use point and line styles to distinguish groups; color is
  optional.
* Add a legend whenever you've used more than one point or line style.
* Always write a few sentences explaining what the plot reveals. Don't
  describe the plot, because the reader can just look at it. Instead,
  explain what they can learn from the plot and point out important details
  that are easily overlooked.
* Sometimes points get plotted on top of each other. This is called
  _overplotting_. Plots with a lot of overplotting can be hard to read and can
  even misrepresent the data by hiding how many points are present. Use a
  two-dimensional density plot or jitter the points to deal with overplotting.
* For side-by-side plots, use the same axis scales for both plots so that
  comparing them is not deceptive.

:::{.callout-note title="See also"}
Visualization design is a deep topic, and whole books have been written about
it. One resource where you can learn more is DataLab's [Principles of Data
Visualization Workshop Reader][dataviz].

[dataviz]: https://ucdavisdatalab.github.io/workshop_data_viz_principles/
:::
